---
title: "eds-223-discussion2"
format: html
---

# Week 2: Discussion Section
## Working with vector operations

### Load the neccessary libraries
```{r}
#| include: false
library(here)
library(tidyverse)
library(sf)
library(tmap)
```

### Read in data (shape files)
Esure a unique coordinate reference system (CRS) is used. Here, we will use NAD 1983 Lambert North America for all spatial datasets.

```{r}
#| include: false
protected_areas <- st_read(here("data","data", "cpad_super_units_sb.shp")) %>% 
  st_transform("ESRI:102009")

city <- st_read(here("data", "data", "sb_city_boundaries_2003.shp")) %>% 
  st_transform("ESRI:102009")

county <- st_read(here("data", "data", "sb_county_boundary_2020.shp")) %>%
  st_transform("ESRI:102009")

birds <- st_read(here("data","data", "aves_observations_2020_2024.shp")) %>% 
  st_transform("ESRI:102009")
```


### Find bird observations within PAs in Santa Barbara County

Find how many bird observations are within protected areas in Santa Barbara County and show how your output differs with a spatial subset versus a spatial join

```{r}
# We can use x[y,] to grab the observations associated with birds in protected_areas 
birds_protected <- protected_areas[birds,]
dim(birds_protected) # There are 35 birds in the protected areas

tm_shape(county) + tm_fill(col = "black") + # basemap 
  tm_shape(protected_areas) + tm_polygons(fill = "tan") +
tm_shape(birds_protected) + tm_dots()
```

```{r}
# Now perform spatial join for both attribute and spatial data
?st_join()

# join protected areas to birds
birds_protected_join <- st_join(birds, protected_areas)

tm_shape(county) + tm_fill(col = "black") + # basemap 
  tm_shape(protected_areas) + tm_polygons(fill = "tan") +
tm_shape(birds_protected_join) + tm_dots()
```

### Find PAs within 15km of Goleta

Find the protected areas within 15 km of a city in Santa Barbara County
Hint: Use dplyr::filter() to select a city from sb_city_boundaries
```{r}
# First, check units
st_crs(protected_areas)$units # Not in km! Need to be careful

# Perform buffer around just goleta by subsetting first
# str(city)
goleta <- city %>% filter(NAME == "Goleta")

# Define a buffer
#?st_buffer()
buffer <- st_buffer(goleta, dist = 15000) # 15*1000

```

Explore different outputs for #3 with st_intersects(), st_intersection(), and st_within()

```{r}
intersection <- st_intersection(protected_areas, buffer)
  
within <- st_within(protected_areas, buffer)

intersect <- st_intersects(protected_areas, buffer)
```


Practice a distance-based join with st_is_within_distance()
```{r}
#?st_is_within_distance()
# We can perform a spatial joinn between protected areas and goleta based on areas almost touching the two sf objects
protected_goleta_distance <- st_join(protected_areas, goleta, fun = st_is_within_distance, dist = 15000)
```


### Find distance between Goleta and Dangermond Preserve
Find the distance between your city of choice and a protected area of your choice, using the geometries’ edges and the centroid.

```{r}
montecito <- city %>% filter(NAME == "Montecito")
#protected_areas$UNIT_NAME
# We have our desired city, but now we also need the preserve
#str(protected_areas)
hot_springs <- protected_areas %>% 
  filter(UNIT_NAME == "Hot Springs Canyon")
```

```{r}
# Find the distance WITHOUT centroid
distance <- st_distance(montecito, hot_springs)

# Find distance WITH centroid (must convert first)
centroid_distance <- st_distance(st_centroid(montecito), st_centroid(hot_springs))
```





