---
title: "eds-223-discussion2"
format: html
---

# Week 2: Discussion Section
## Working with vector operations

### Load the neccessary libraries
```{r}
#| include: false
library(here)
library(tidyverse)
library(sf)
library(tmap)
```

### Read in data (shape files)
Esure a unique coordinate reference system (CRS) is used. Here, we will use NAD 1983 Lambert North America for all spatial datasets.

```{r}
#| include: false
protected_areas <- st_read(here("data","data", "cpad_super_units_sb.shp")) %>% 
  st_transform("ESRI:102009")

city <- st_read(here("data", "data", "sb_city_boundaries_2003.shp")) %>% 
  st_transform("ESRI:102009")

county <- st_read(here("data", "data", "sb_county_boundary_2020.shp")) %>%
  st_transform("ESRI:102009")

birds <- st_read(here("data","data", "aves_observations_2020_2024.shp")) %>% 
  st_transform("ESRI:102009")
```


### Find bird observations within PAs in Santa Barbara County

Find how many bird observations are within protected areas in Santa Barbara County and show how your output differs with a spatial subset versus a spatial join

```{r}
# We can use x[y,] to grab the observations associated with birds in protected_areas 
birds_protected <- protected_areas[birds,]
dim(birds_protected) # There are 35 birds in the protected areas

tm_shape(county) + tm_fill(col = "black") + # basemap 
  tm_shape(protected_areas) + tm_polygons(fill = "tan") +
tm_shape(birds_protected) + tm_dots()
```

```{r}
# Now perform spatial join for both attribute and spatial data
?st_join()

# join protected areas to birds
birds_protected_join <- st_join(birds, protected_areas)

tm_shape(county) + tm_fill(col = "black") + # basemap 
  tm_shape(protected_areas) + tm_polygons(fill = "tan") +
tm_shape(birds_protected_join) + tm_dots()
```






